# Reproducible Research: Peer Assessment 1
|                |                                                        |
|:---------------|:-------------------------------------------------------|
|**Author:**     | Brian von Konsky
|**Created:**    | May 2014
|**Course:**     | Reproducible Research
|**Repository:** | [https://github.com/bvonkonsky/RepData_PeerAssessment1](https://github.com/bvonkonsky/RepData_PeerAssessment1)

## Loading and preprocessing the data

**Step 1.** Specify the location of files and directories.
```{r}
# Set file names
original  <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
directory <- "."
zip       <- "activity.zip"
data      <- "activity.csv"
        
# Create OS idependent filenames for path to zip archive and data
path.zip <- file.path(directory, zip)
path.data<- file.path(directory, data)
```
**Step 2.** Download the zip archive if it doesn't exist in the current working
directory.
```{r}
if (!file.exists(zip)) {    
        download.file(original, path.zip, method="curl" )
        }
        
```
**Step 3.** Unzip the archive and read the data.
```{r}
# Unzip the archive
unzip(path.zip)
        
# Read the data from the csv file
activity <-read.csv(path.data, header=TRUE)
```
**Step 4.** Convert dates to objects of class Date.
```{r}
activity$date <- as.Date(activity$date)
```
**Step 5.** Subset and aggregate data that will be used in this analysis. 
As shown on the [aggregate](http://stat.ethz.ch/R-manual/R-devel/library/stats/html/aggregate.html) help page, the **na.action** default is to ignore NA
values.
```{r}
# Aggregate to caluclate the total number of steps per day
stepsPerDayNoNA  <- aggregate(steps ~ date, activity, sum)
names(stepsPerDayNoNA) <- c("date", "steps.sum")

# Aggregate to calculate the average number of steps per interval
avgPerIntervalNoNA <- aggregate(steps ~ interval, activity, mean)
names(avgPerIntervalNoNA) <- c("interval", "steps.mean")
```
**Step 6.** Inspect the full data frame and its subsets to confirm correct 
processing.
```{r}
str(activity)
str(stepsPerDayNoNA)
str(avgPerIntervalNoNA)
```

## What is mean total number of steps taken per day?

A histogram showing the frequency with which the subject
walked a given total steps per day is shown below. The mean number of steps per day is shown with a blue line.
```{r, fig.width=5, fig.height=5, fig.align="center"}
totalSteps <- function(stepsPerDay, 
                                  title="Frequency of Total Steps per Day", 
                                  yrange=c(0,12)) {
        with (stepsPerDay, {
                
                # Plot the frequency of Total Steps per Day
                hist(steps.sum,
                     col="red",
                     breaks=20,
                     xlab="Total Steps per Day",
                     main=title,
                     xlim=c(0, 25000), ylim=yrange
                )
                
                # Use a blue line to show the mean value
                abline(v=mean(steps.sum), col="blue", lwd=3)
                })
              
}

# generate a histogram for the total number of steps per day 
totalSteps(stepsPerDayNoNA)
```
The mean total number of steps per day is  **10766.19**. This was calculated by
calling the `summary()` function with the significant **digits** parameter set
to 7. The number of significant **digits** was set in the call to `summary()` so 
that that it was consistent with the equivalent value returned by `mean()`.
```{r}
with (stepsPerDayNoNA, summary(steps.sum, digits=7))
```

## What is the average daily activity pattern?

A plot showing the time series with the number steps per 5 minute interval,
averaged across all days is shown below. The interval with the maximum average
number of steps per interval is shown with a blue line.  The maximum average
steps per interval and the interval in which it occurs is also calculated in 
[R](http://www.r-project.org/). The
maximum value is **206.1698**, and occurs in 5 minute interval number **835** in
row **104** of the aggregated data.
```{r, fig.width=5, fig.height=5, fig.align="center"}
with (avgPerIntervalNoNA, {
        # Determine which row contains the maximum avergage 
        # number of steps per interval
        maxRow <- which.max(steps.mean)
                
        # Plot the time series
        plot(interval, steps.mean, type="l",
             xlab="Interval Number", ylab="Average Number of Steps",
             main="Average Number of Steps per 5 Minute Inerval")  
                
        # Show where the maximum average steps occur
        abline(v=interval[maxRow], col="blue")
                
        # Display the maximum value and the interval in which it occurs
        print(avgPerIntervalNoNA[maxRow,])
        })
```

## Inputing missing values
The number of observations containing NA values for the number of steps, and
the number of days on which this occurs is calculated below.
```{r}
activity.na <- subset(activity, is.na(steps))
days   <- unique(activity.na$date)
num.na <- nrow(activity.na)
num.days  <- length(days)
print(paste("There are", num.na, "NA entries on", 
                    num.days, "unique days"))
```
The 8 days containing NA values are given below.
```{r}
days
```
NA values were replaced by the mean value for the intervals in which they occurred. Given that there were 8 days in which all values were NA, this strategy had the net impact of increasing the frequency of total steps per day at the median by 8. This is demonstrated in the plot below.
```{r, fig.width=10, fig.height=5, fig.align="center"}
activityAvgNA <- activity
for (i in c(0, seq(5,2355,5))) {
        match <- is.na(activityAvgNA$steps) & (activityAvgNA$interval==i)
        activityAvgNA$steps[match] <- 
                mean(activityAvgNA$steps[activityAvgNA$interval==i], 
                     na.rm=TRUE)
}

stepsPerDayAvgNA   <- aggregate(steps ~ date, activityAvgNA, sum)
names(stepsPerDayAvgNA) <- c("date", "steps.sum")

par(mfrow=c(1,2))
totalSteps(stepsPerDayNoNA, "NA Removed", yrange=c(0,20))
totalSteps(stepsPerDayAvgNA,"NA Replaced by Interval Average", yrange=c(0,20))
```
As expected, the mean and median values remained the same, despite the increase
in the frequency with which the mean total steps per day occurred.
```{r}
summary(stepsPerDayNoNA$steps.sum)
summary(stepsPerDayAvgNA$steps.sum)
```

## Are there differences in activity patterns between weekdays and weekends?
The subject was the most active early in the morning on weekdays, but overall
was generally more active towards the middle of the day on weekends.
```{r, fig.width=8, fig.height=10, fig.align="center"}

# Generate days of the week for activity in which NA values have been
# replaced with the average for that interval
dow <- weekdays(activityAvgNA$date, abbreviate=FALSE)

# Generate factors for weekend versus weekdays
fac <- rep("unset", length(dow))
fac[dow=="Sunday" | dow=="Saturday"] <- "weekend"
fac[dow!="Sunday" & dow!="Saturday"] <- "weekday"

# Add the factors to the data
activityAvgNA <- cbind(activityAvgNA, day.week=dow, day.type=fac)
names(activityAvgNA) <- c("steps", "date", "interval", "day.week", "day.type")

# Subset the data by weekend and weekdays
weekday <- subset(activityAvgNA, day.type=="weekday")
weekend <- subset(activityAvgNA, day.type=="weekend")

# Compute the mean for aggregated weekday data
weekdayAvgPerInterval <- aggregate(steps~interval, weekday, mean)
names(weekdayAvgPerInterval) <- c("interval", "steps.mean")

# Compute the mean for aggregated weekend data
weekendAvgPerInterval <- aggregate(steps~interval, weekend, mean)
names(weekendAvgPerInterval) <- c("interval", "steps.mean")

# Plot the result
par(mfrow=c(2,1), mar=c(2,3,1,1),  pin=c(7, 3))
with(weekdayAvgPerInterval, plot(interval, steps.mean, type="l", yaxp=c(0, 200,2), ylim=c(0,200), xlab="", ylab="Average Steps", main="weekday"))
with(weekendAvgPerInterval, plot(interval, steps.mean, type="l", yaxp=c(0,200, 2), ylim=c(0,200), ylab="Average Steps", main="weekend"))
```
